Accurate interactions of cations with neutral and negatively charged membranes via combination of experiments and molecular simulation

### 2018-08-06 Manuscript/manuscript.tex
Small update, comments.
Major update to the manuscript:   copied-in contents of my thesis for further editing.   Updated packages in latex template   updated biliography   plots are updated (my doctoral thesis submittion state)

\usepackage{rotating}  
\usepackage{times} 
\usepackage{graphicx} 
\usepackage{setspace} 
\usepackage{amsmath} 
\usepackage{epstopdf} 
\usepackage{csquotes} 
\usepackage{mhchem} 
\usepackage{chemfig}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%% Place any additional macros here.  Please use \newcommand* where 
%% possible, and avoid layout-changing macros (which are not used 
%% when typesetting). 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\newcommand*\mycommand[1]{\texttt{\emph{#1}}} 
\newlength{\figwidth}
\setlength{\figwidth}{8 cm}

@@

\begin{abstract} 
Binding affinities and stoichiometries of sodium and calcium ions to phospholipid bilayers are of paramount significance in the properties and functionality of cellular membranes. Current estimates of binding affinities and stoichiometries of cations are, however, controversial due to limitations in the available experimental and theoretical methods. Experimentally one can assess these parameters by titrating several membrane properties as a function of the ion concentration. However, the interpretation of the experiments relies on theoretical models, as direct molecular information is not available. 
Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic resolution, therefore offering all the necessary information to interpret experimental data without the need to resort to over-simplified models. However, the accuracy of the currently available lipid models when interacting with ions is not sufficient for such interpretation \cite{catte16}. Recently, we have improved the binding details of sodium and calcium ions to the 1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic polarization as a mean field correction, known as the electronic continuum correction (ECC). Our improved ECC-POPC model reproduces not only the experimentally measured structural parameters for the ion-free membrane, but also improves significantly the response of lipid head group to a bound positive charge, and the binding affinities of sodium and calcium ions. \cite{melcr18}
We are now extending our work to other lipid membranes composed of larger varieties of different biologically relevant lipids. In particular, we focus our attention on membranes with phosphatidylethanolamine (PE) and negatively charged phosphatidylserine (PS). We compare available NMR and x-ray experiments with MD simulations to obtain atomistic-resolution picture of the interactions between cations and phospholipid membranes. Our preliminary results indicate that electronic polarization is a non-negligable interaction, which dramatically improves the description of cation binding to the membranes. 
\end{abstract}

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc} 
This section may not be required, we may just cite ECC-POPS 
and put a short paragraph describing what we did on POPS.

Compared to ECC-POPC model \cite{melcr18}, 
which used used a scaling factor $f_q = 0.8$ for all charges but lipid tails,
this model, ECC-POPS, has to adopt a scaling factor $f_q = 0.75$ 
as it has non-zero total charge in line with ECC theory. \cite{leontyev09}
Hence, electronic continuum correction of POPS 
has the same characteristics as for cations and anions, 
denoted as ECC-ions \cite{martinek17, Pluharova2014, kohagen14, kohagen16}.

@@

\subsubsection{Simulations of POPS:POPS bilayers with aqueous ions}

Any specifics for simulations with cations.

\subsubsection{Simulations of POPS:POPS bilayers with anionic/cationic surfactants/peptides}

@@

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_actual_pure-POPS.pdf} \\
  \includegraphics[width=0.7\figwidth]{../Fig/form-f_l17-ecc-pops-exp_compar.pdf} 
  \includegraphics[width=0.55 \figwidth]{../Fig/lipids_chemfig.pdf} 
  \caption{\label{simVSexpNOions} 
    X-ray scattering form factors from simulations with the Lipid17 \cite{lipid17-future} and 
    the ECC-POPS models compared with experiments~\cite{SDP-CHARMM36_comparison_paper_Samuli-knows} at 298~K. 
    The disagreements between ECC-POPS model and experiment 
    may come from an intrinsincally incorrect structure of the head group (and, hence, electron density in the region), 
    however, lipid packing and overall membrane properties 
    may be still caputured well enough -- after considering areas per lipid reported in Table~\ref{tab:apls}.  \\
    Order parameters of POPS head group, glycerol backbone and acyl chains  
    from simulations with the Lipid17 \cite{lipid17-future} and the ECC-POPS models 
    compared with \emph{experiments at 300~K (check and change)}. 
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPS. \\
    The chemical structure of POPS and the labeling of the carbon segments. 
  }  
\end{figure}

\begin{table}[tb!] 
  \caption{Values of the area per lipid (APL) of POPS bilayers with only \ce{Na^+} counterions. \label{tab:apls} 
  } 
  \begin{tabular}{l|c c} 
    model          & APL (\AA$^2$)   & Temperature [K] \\ 
    \hline 
    Lipid14                   & 53.5$\pm$ 0.8  &  298 \\ 
    \hline 
    ECC-POPS                & 60.3$\pm$ 0.6  &  298       \\ 
    \hline 
    experiment \cite{SDP-CHARMM36_comparison_paper_Samuli-knows} & 62.3  &  298    \\ 
    \hline 
  \end{tabular} 
\end{table}

\subsection{Electrometer concept in negatively charged membranes with various cations}

Preliminary results are plotted in the figures in this section. 
Beware! -- the y-axis scales are variable!
Colours:
\begin{itemize}
  \item Red -- ECC-lipids, ECC-ions
  \item Blue -- Lipid14/17, Dang ions
\end{itemize}

Changes of the lipid bilayer head group order parameters extracted from simulations and 
experiments \cite{roux90} are shown in Figs.~\ref{fig:delta_ordPar_KCl}, \ref{fig:delta_ordPar_NaCl} 
and~\ref{fig:delta_ordPar_CaCl} as functions of KCl, NaCl or CaCl$_2$ concentrations,
and in Figs.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix} as a function of PC and PS content.

% ============================== 
% ==  ECC-lipids plots ========= 
% ==     and           ========= 
% ==  Lipid14/17 plots ========= 
% ============================== 
\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/l17/order_parameters_changes_A-B_POPC_kcl.pdf} 
  \includegraphics[width=\figwidth]{../Fig/l17/order_parameters_changes_A-B_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl_l17} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of KCl concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K together with  
  } 
\end{figure}

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_changes_A-B_POPC_kcl.pdf} 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_changes_A-B_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of KCl concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K together with  
  } 
\end{figure}

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/l17/order_parameters_changes_A-B_POPC_nacl.pdf} 
  \includegraphics[width=\figwidth]{../Fig/l17/order_parameters_changes_A-B_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_l17} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of NaCl concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K together with  
  } 
\end{figure}

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_changes_A-B_POPC_nacl.pdf} 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_changes_A-B_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of NaCl concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K together with  
  } 
\end{figure}

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/l17/order_parameters_changes_A-B_POPC_cacl.pdf} 
  \includegraphics[width=\figwidth]{../Fig/l17/order_parameters_changes_A-B_POPS_cacl.pdf} 
  \caption{\label{fig:delta_ordPar_CaCl_l17} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of CaCl concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K together with  
  } 
\end{figure}

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_changes_A-B_POPC_cacl.pdf} 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_changes_A-B_POPS_cacl.pdf} 
  \caption{\label{fig:delta_ordPar_CaCl} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of CaCl concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K together with  
  } 
\end{figure}

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/l17/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[width=\figwidth]{../Fig/l17/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix_l17} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of PS content.
    Experimental points have inherent systematic error, as they come from different experiments.
    A complete series of points can be found (?) in \cite{roux90} -- which is to be added to the plot.
  } 
\end{figure}

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[width=\figwidth]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of PS content
    Experimental points have inherent systematic error, as they come from different experiments.
    A complete series of points can be found (?) in \cite{roux90} -- which is to be added to the plot.
  } 
\end{figure}

\subsection{Molecular interaction and binding affinities of K$^+$, Na$^+$ or Ca$^{2+}$  cations to the mixed POPC:POPS (5:1) membrane} 
\label{sec:affinity}

>>>>

%% Workaround for malfunctioning textsuperscript with pdfx and T1 encoding
\let\tmpa\textsuperscript
\DeclareTextCommandDefault{\textsuperscript}{\tmpa}

\usepackage[x11names]{xcolor}  % typesetting in color
%% Generate PDF/A-2u
\usepackage[a-2u]{pdfx}

\usepackage[super]{natbib}         % citation style AUTHOR (YEAR), or AUTHOR [NUMBER]

\usepackage{achemso}  
\usepackage{rotating}  
%\usepackage{times} 
%\usepackage{lmodern}    %% Prefer Latin Modern fonts
\usepackage{graphicx} 
\usepackage{setspace} 
\usepackage{amsmath} 
\usepackage{epstopdf} 
\usepackage{csquotes} 
\usepackage{mhchem} 
\usepackage{chemfig} 
\usepackage[obeyFinal]{easy-todo}
\usepackage[english]{babel}

\usepackage{subfig}
\usepackage{dcolumn}        % improved alignment of table columns
\usepackage{booktabs}       % improved horizontal lines in tables
\usepackage{paralist}       % improved enumerate and itemize

%% Character encoding: usually latin2, cp1250 or utf8:
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

@@

%% The hyperref package for clickable links in PDF and also for storing
%% metadata to PDF (including the table of contents).
%% Most settings are pre-set by the pdfx package.
\hypersetup{unicode}
\hypersetup{breaklinks=true}
\hypersetup{colorlinks,allcolors=DodgerBlue4}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%% Place any additional macros here.  Please use \newcommand* where 
%% possible, and avoid layout-changing macros (which are not used 
%% when typesetting). 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\newcommand*\mycommand[1]{\texttt{\emph{#1}}} 
%%% Custom variables
% width suitable for fitting into a column in 1-column page layout
\newlength{\figwidth}
\setlength{\figwidth}{9 cm} 
\newlength{\figwidthsmall}
\setlength{\figwidthsmall}{6 cm} 
\newlength{\figwidthfull}
\setlength{\figwidthfull}{14 cm}

@@

\begin{abstract} 
	\emph{Adapt abstract to the final message: properties of POPS-containing bilayers -or- ECC-concept applicability?}

Binding affinities and stoichiometries of sodium and calcium ions to phospholipid bilayers are of paramount significance in the properties and functionality of cellular membranes. Current estimates of binding affinities and stoichiometries of cations are, however, controversial due to limitations in the available experimental and theoretical methods. Experimentally one can assess these parameters by titrating several membrane properties as a function of the ion concentration. However, the interpretation of the experiments relies on theoretical models, as direct molecular information is not available.

Classical molecular dynamics (MD) simulations provide details of the ion binding process with atomistic resolution, therefore offering all the necessary information to interpret experimental data without the need to resort to over-simplified models. However, the accuracy of the currently available lipid models when interacting with ions is not sufficient for such interpretation \cite{catte16}. Recently, we have improved the binding details of sodium and calcium ions to the 1-Palmitoyl-2-oleoyl-phosphatidylcholine (POPC) bilayer by implicitly including electronic polarization as a mean field correction, known as the electronic continuum correction (ECC). Our improved ECC-POPC model reproduces not only the experimentally measured structural parameters for the ion-free membrane, but also improves significantly the response of lipid head group to a bound positive charge, and the binding affinities of sodium and calcium ions. \cite{melcr18}

We are now extending our work to other lipid membranes composed of larger varieties of different biologically relevant lipids. In particular, we focus our attention on membranes with 
	%phosphatidylethanolamine (PE) and 
	negatively charged phosphatidylserine (PS) lipids. 
	We compare available NMR and x-ray experiments with MD simulations to obtain atomistic-resolution picture of the interactions between cations and phospholipid membranes. Our results indicate that electronic polarization is a non-negligable interaction, which dramatically improves the description of cation binding to phospholipid membranes. 
\end{abstract}

@@

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}

\emph{ Directly taken from my thesis.}

The lack of electronic polarizability in classical MD simulations
has been considered a serious issue since the early days of the modeling of lipid bilayers. 
% the commented text on polarizability in this section comes from \citep{huang2017mapping}
There exist two main approaches for the explicit representation of electronic polarizability -- atomic induced dipoles and Drude oscillators.
%The AMOEBA FF adopts the point induced dipole model in which each atom carries an induced dipole moment determined by the electric field and its atomic polarizability. 
%Permanent dipole and quadrupoles are also assigned to each atomic site for a more accurate description of electrostatic interactions. 
The induced dipole approach models the polarizability through a point dipole moment on each atom.
Such an electric dipole is induced by the acting electric field proportionally to its polarizability. 
This model is implemented in the widely adopted force field AMOEBA \citep{amoeba06, amoeba10, amoeba13}.

%The Drude FF adopts the classical Drude oscillator model where auxiliary charges (Drude oscillators or particles) are attached to non-hydrogen atoms with harmonic springs, and the displacement of the Drude particle relative to its parent atom represents the response to the external electric field. 
%Virtual particles representative of lone pairs are included in the Drude FF to improve the treatment of electrostatic interactions typically involving hydrogen bond acceptors.
%The charge distribution between a Drude particle and its parent atom is a finite difference representation of a dipole moment. 
%Similarly, lone pairs can be considered as real space representations of permanent multipoles. 
%Such analogy between the classical Drude oscillator model and the multipole and induced dipole (MPID) model suggests that in theory one can interconvert polarizable FFs between the two models. 
The classical Drude oscillator model is inspired by a classical representation of a lone pair of electrons. 
It uses an auxiliary charge,
also termed Drude oscillator or Drude particle, 
which is attached to each atom through a harmonic spring.
The displacement of the Drude particle represents the induced polarization proportionally to the acting electric field,
as implemented for example in the Drude force field \citep{lemkul2016empirical}.

%In this work, we convert the Drude polarizable force field from its original classical Drude oscillator formalism onto a permanent multipole and induced dipole formalism, which we refer to as MPID. 
%We show that a direct mapping without any reparametrization can transfer the Drude FF into the MPID FF, yielding a similar level of agreement as the Drude FF in reproducing condensed phase properties of organic liquids, crystals, and interfaces. 
%Our results thus also serve as an additional validation of these novel simulation algorithms.
Although the two approaches, 
the induced dipole model and the classical Drude oscillator model, 
are very different in how they represent polarizability in simulations,
it was shown that the classical Drude oscillator formalism can be converted to the induced dipole formalism 
proving the two methods to be equivalent. \citep{huang2017mapping}
From the point of view of lipids, however, the classical models with fixed charges and no explicit polarizability that are currently available 
provide comparable accuracy at a fraction of the computational cost of the models with explicit polarization. \citep{lucas12,chowdhary13}

%The effects of polarization from rather computationally demanding simulations with explicit models 
%can be implemented using an implicit model of polarization. 
Instead of the computationally demanding explicit polarization approach, 
electronic polarization can also be accounted for implicitly.
Such an implicit mean field model of electronic polarizability 
%was developed by \citet{leontyev09} under the name Molecular Dynamics in Electronic Continuum, MDEC. 
termed Electronic continuum correction (ECC) 
was first suggested by \citet{leontyev09}, 
and was then systematically applied to solutions of ions by \citet{Pluharova2014, kohagen14, kohagen16, martinek17}. 
Through a comparison of neutron scattering and/or \emph{ab-initio} molecular dynamics data with classical MD simulations, 
the authors developed an array of models for cations and anions, termed here as ECC-ions,
which provide realistic structures even for concentrated solutions of both monovalent and divalent cations. 
%The term Electronic continuum correction, ECC, originates in the works by \citet{Pluharova2014, kohagen14, kohagen16, martinek17},
%where the rigorously developed theory by \citet{leontyev14} was systematicaly applied to solutions of ions.
%that use ECC to represent electronic polarizability.

%Electronic continuum correction (ECC) 
In ECC, a system of polarizable particles is represented 
with an equivalent system of particles with fixed charges. %\citep{leontyev09, leontyev10, leontyev11, leontyev14}
%including the effects of electronic polarization \emph{implicitly} in a mean-field way. 
This can be done by a simple transform of the partial charges,
which does not change the form of the empirical force field in Equation~\ref{eq:amber}, 
providing the effects of the electronic polarization in a mean field way at no additional computational cost. 
Although this is technically similar to the phenomenological scaling of partial charges 
applied in earlier studies of aqueous solutions or ionic liquids \citep{jonsson86,egberts94,beichel14},
%where also other possible effects (e.g. charge transfer) were considered,
ECC is a physically well justified and rigorously derived model \citep{leontyev09, leontyev10, leontyev11, leontyev14}.
Assuming that the electronic polarizability of electrons is approximately constant in biological solutions, 
it was shown by \citet{leontyev11} that the electrostatic interactions between atoms 
are screened like in a dielectric continuum 
with an electronic dielectric constant, $\epsilon _{el}$, 
also known as the high frequency dielectric constant. 
This can be formally written as
\begin{equation}  \label{eq:coulomb}
   %U = \frac{1}{2} \displaystyle \sum ^N _{j\neq i} \frac{q_i q_j}{r_{ij}} + \frac{1}{2} \displaystyle \sum ^N _{i,j=1} U ^{dip} _{ij} - \displaystyle \sum ^N _{i=1} E^q(r_i) d_i  \propto  \frac{1}{2} \displaystyle \sum ^N _{j\neq i} \frac{q_i q_j}{\epsilon_{el} \, r_{ij}}  =  \frac{1}{2} \displaystyle \sum ^N _{j\neq i} \frac{q^{ECC}_i q^{ECC}_j}{r_{ij}} 
   %U = \frac{1}{2} \displaystyle \sum ^N _{j\neq i} \frac{q_i q_j}{r_{ij}} + \frac{1}{2} \displaystyle \sum ^N _{i,j=1} U ^{dip-dip} _{ij} - \displaystyle \sum ^N _{i=1} U^{q-dip}_i  \propto  \frac{1}{2} \displaystyle \sum ^N _{j\neq i} \frac{q_i q_j}{\epsilon_{el} \, r_{ij}}  =  \frac{1}{2} \displaystyle \sum ^N _{j\neq i} \frac{\frac{q_i}{\sqrt{\epsilon _{el}}} \frac{q_j}{\sqrt{\epsilon _{el}}}}{r_{ij}}  , 
   U = \frac{1}{2} \displaystyle \sum ^N _{j\neq i} \frac{q_i q_j}{\epsilon_{el} \, r_{ij}}  =  \frac{1}{2} \displaystyle \sum ^N _{j\neq i} \frac{\frac{q_i}{\sqrt{\epsilon _{el}}} \frac{q_j}{\sqrt{\epsilon _{el}}}}{r_{ij}}  , 
\end{equation} 
where $q_i$ is the point charge of the particle $i$,
and $r_{ij}$ are the distances between particles $i$ and $j$.
%$U^{dip-dip} _{ij}$ represent the interactions between induced dipoles and their polarization energies,
%and the terms $U^{q-dip} _i$ are the interactions between point charges and induced dipoles. 
It naturally follows from the above equation that 
embedding all atoms into a dielectric continuum 
is equivalent to a formal scaling of the atomic point charges 
\begin{equation}  \label{eq:scaling}
 q_i ^{ECC} = \frac{q_i}{\sqrt{\epsilon _{el}}} ,
\end{equation} 
where $q_i$ and $q_i^{ECC}$ are respectively the original partial charges and the scaled charges, 
which represent the effects of the electronic polarization. 
Given that the  high frequency dielectric constant of water is 1.78 (i.e., the square of the refraction index of 1.33), 
the scaling factor for ions in water is $\approx 0.75$.

It is important to note that the value of the high frequency dielectric constant  
is around 2 for almost any biologically relevant environment \citep{leontyev11}. 
This means that even interfaces like biological membranes do not contain discontinuities of the electronic continuum. 
The dielectric discontinuity in a lipid bilayer thus arises only 
from the orientational polarization of the molecules, which is accounted for explicitly in standard MD simulations.  
Therefore, the same correction for the electronic polarizability can be  
applied throughout the lipid bilayer/aqueous solution interface.

@@

Any specifics for simulations with aqueous cations.

@@

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/Order-parameters_form-factors_exp-L17-ECC-lipids.pdf}
  \includegraphics[width=\figwidth]{../img/ecc_pops/pops_chemfig.pdf} 
\hfill
  \caption{\label{simVSexpNOions_POPS} 
    Top: X-ray scattering form factors from simulations with the Lipid17 \citep{lipid17-future} and 
    the ECC-POPS models compared with experiments~\citep{kucerka14} at 298~K. 
    Middle: Order parameters of POPS head group, glycerol backbone and acyl chains  
    from simulations with the Lipid17 \citep{lipid17-future} and the ECC-POPS models 
    compared with experiments at 298~K. \citep{nmrlipids_proj4}
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPS. 
    Bottom: The chemical structure of POPS and the labeling of the carbon segments. 
  }  
\end{figure}

\begin{table}[tb!] 
\centering
  \caption{Values of the area per lipid (APL) of POPS bilayers 
	   at 298~K with only \ce{Na^+} counterions and no additional ions. \label{tab:apls} } 
  \begin{tabular}{l|c c} 
%    \multicolumn{3}{c}{POPC} \\
%    model          & APL / Å$^2$   & $T$ / K  \\ 
%    \hline 
%    Lipid14 POPC \citep{melcr18}    & 65.1$\pm$ 0.6  &  300 \\ 
%    Lipid14 POPC \citep{dickson14}  & 65.6$\pm$ 0.5  &  303 \\ 
%    \hline 
%    ECC-POPC     \citep{melcr18}    & 63.2$\pm$ 0.6  &  300       \\ 
%    \hline 
%    experiment (SDP model) \citep{kucerka11} & 64.3  &  303    \\ 
%    \hline 
%    \multicolumn{3}{c}{} \\
%    \multicolumn{3}{c}{POPE} \\
%    model          & APL / Å$^2$   & $T$ / K  \\ 
%    \hline 
%    ECC-POPE                 & 56.7$\pm$ 0.8  &  298 \\ 
%    \hline 
%    experiment   \citep{parsegian89} & 56.6  &  310    \\ 
%    experiment   \citep{rappolt03}   & 59--61 &  303--313  \\ 
%    \hline 
%    \multicolumn{3}{c}{} \\
    \multicolumn{3}{c}{POPS} \\
    model          & APL / Å$^2$   & $T$ / K  \\ 
    \hline 
    Lipid17 POPS              & 53.5$\pm$ 0.8  &  298 \\ 
    \hline 
    ECC-POPS                & 60.3$\pm$ 0.6  &  298       \\ 
    \hline 
    experiment (SDP model) \citep{kucerka14} & 62.3  &  298    \\ 
    \hline 
  \end{tabular} 
\end{table}

We compared X-ray scattering form factors and NMR order parameters of bilayers
in pure water without any ions (or only counter ions)
from simulations and experiments
as the first step in the assessment of the quality of the model. 
The experimental X-ray scattering form factors 
of a bilayer are well reproduced for all lipids employing the presented ECC-lipids model 
(see Figs.~\ref{fig:simVSexpNOions}, \ref{simVSexpNOions_POPE} and \ref{simVSexpNOions_POPS}).

The area per lipid is often used as a relatively simple structural parameter reporting on the bilayer properties and the packing of lipids. 
In experiments, modeling is used on top of the scattering factors to obtain it \citep{kucerka14}. 
From simulations, this property is easily extracted. 
We compare the values from experiments and simulations in Table~\ref{tab:apls}.

While the agreement between the scattering form factors 
from the simulation of a pure POPS bilayer and experiment 
are excellent (Fig.~\ref{simVSexpNOions_POPS}),
there is a non-negligible difference between the values of the area per lipid in Table~\ref{tab:apls}. 
Since both values are derived from the scattering form factors through modeling of the electron density of the bilayer,
we cannot decide, which of the values is more reliable. 
In general, we can conclude that ECC-lipids
%the presented lipid models with ECC-lipids 
reproduce the experimental structural parameters of the lipid bilayers 
with a comparable accuracy to existing state-of-the-art lipid models~\citep{botan15, ollila16, Pluhackova2016}.

The head group and acyl chain order parameters within ECC-lipids
are in general in a good agreement with the experimental values 
as shown in Fig. \ref{simVSexpNOions_POPS}. 
The acyl chain order parameters in particular are almost all within the experimental error bars.
The order parameters of the head groups are at an accuracy comparable to 
other currently available classical models of lipids \citep{botan15, catte16, Pluhackova2016}.

The head group order parameters $\alpha$ and $\beta$ are highly relevant for this work,
as they are being used in the electrometer concept \cite{altenbach84, catte16, melcr18}.
For POPC in pure water, the order parameter $\beta$ agrees well with the experiment, 
while the order parameter $\alpha$ is somewhat lower. 
In the case of POPS, the situation is a bit more complicated
compared to POPC as the order parameter $\alpha$ exhibits a notable forking (see Fig.~\ref{simVSexpNOions_POPS}).
One of the order parameters of ECC-POPS, $\alpha_1$, agrees well with the experiment, 
while the other, $\alpha_2$, adopts a higher value underestimating the experimentally reported forking. 
There is only one order parameter $\beta$ in POPS, 
which has a higher value closer to zero in the ECC-lipids model than in experiment. 
Such a feature suggests that the model overestimates the orientational freedom of its head group.

\subsection{Electrometer concept: cationic and anionic amphiphilic molecules}

Changes of the lipid bilayer head group order parameters extracted from simulations and 
experiments \cite{roux90} are shown in Figs.~\ref{fig:delta_ordPar_KCl_PCPS}, \ref{fig:delta_ordPar_NaCl_PCPS} 
and~\ref{fig:delta_ordPar_CaC_PCPS} as functions of KCl, NaCl or CaCl$_2$ concentrations,
and in Figs.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix} as a function of PC and PS content. 
and in Figs.~\ref{OrderParameterCHANGESsurf} as a function of the content of dihexadecylammonium bromide, a cationic amphiphile.

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_popc/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf} 
  \caption{\label{OrderParameterCHANGESsurf} 
    Changes of the head group order parameters $\alpha$, $\beta$ and P-N vector orientation as a function of 
    the molar fraction of a cationic surfactant dihexadecyldimethylammonium in a POPC bilayer 
    from simulations and experiments \citep{scherer89} at 313 K.
    Image reproduced from \cite{melcr18}.
  } 
\todo{Combine with Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix} into a single plot.}
\end{figure}

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../Fig/l17/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/l17/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix} 
    Changes of the head group order parameters of a POPC:POPS 5:1 bilayer as a function of PS content
    Experimental points have inherent systematic error, as they come from different experiments.
    A complete series of points can be found (?) in \cite{roux90} -- which is to be added to the plot.
  } 
\end{figure}

\subsection{Molecular interaction and binding affinities of K$^+$ and Na$^+$ cations to mixed POPC:POPS (5:1) membrane} 
\label{sec:affinity}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer as a function of \ce{NaCl} concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K.
    Because data with \ce{NaCl} are not available for POPC, 
    we show experimental data for \ce{LiCl} (dashed line, left) 
    as an upper bound for the magnitude of the response to \ce{NaCl}, 
    which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  } 
\end{figure}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_kcl.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer as a function of \ce{KCl} concentration 
    in bulk ($C_{ion}$) from simulations with different force fields and experiments at 298 K. \citep{roux90}
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  } 
\end{figure}

% Pavel suggested to remove this as it only recaps previous sections
%It was presented in the previous section,
%that accounting for electronic polarization,
%a distinct feature of ECC-lipids compared to other lipid models, 
%is crucial for an accurate description of the response of the POPC electrometer. 
%In this section,
%we employ ECC-lipids to simulate neutral and negatively charged bilayers 
%in the presence of monovalent salts, namely \ce{KCl} and \ce{NaCl}.

The binding of \ce{Na^+} cations to phospholipid bilayers 
is not generally agreed on between simulations 
\citep{bockmann03,sachs04,berkowitz06,cordomi09}
and experiments 
\citep{cevc90,tocanne90,hauser76,herbette84,uhrikova08}.
We attempted to find a model, 
which would successfully interpret the experiments on binding of \ce{Na^+}
to neutral and negatively charged phospholipid bilayers \citep{akutsu81, roux90} 
in our works \citep{catte16, nmrlipids_proj4}.
After testing the currently available force fields for phospholipids,
we concluded
that the interactions are generally overestimated in magnitude in almost all models 
but Lipid14 (PC) \citep{dickson14}, resp. Lipid17 (PS) \citep{lipid17-future}. 
This model yields a semi-quantitative agreement with the experimentally measured small changes of the order parameters 
when used with the model of ions by \citet{aqvist90} (Fig.~\ref{fig:catte16}). 
However, when used with a more accurate model of ions by \citet{Pluharova2014, martinek17},
the model overestimates the binding affinity of \ce{Na^+}
measured with lipid electrometer concept. \citep{melcr18}
In total, these results suggest that improvements 
in the lipid parameters are required for more accurate interactions even with monovalent cations. 
\citep{catte16, melcr18, nmrlipids_proj4}

Generally improved behaviour 
of the POPC and POPS head group order parameters 
with \ce{NaCl} or \ce{KCl} concentrations 
was achieved through the combination of models 
ECC-lipids \citep{melcr18} and ECC-ions \citep{martinek17, kohagen16, Pluharova2014}. 
Simulations with these models 
reveal a good agreement with the NMR experiments 
for both neutral and negatively charged membranes.
The results are plotted for \ce{NaCl} in Fig. \ref{fig:delta_ordPar_NaCl_PCPS}, 
and for \ce{KCl} in Fig. \ref{fig:delta_ordPar_KCl_PCPS}.
